---
title: "Constructing cell-type-specific networks in human prefrontal cortex"
output: html_notebook
---

## Import libraries
```{r message=FALSE, warning=FALSE}
library(Matrix)
library(ACTION)
library(igraph)
library(RColorBrewer)
library(ggplot2)
```


## Process expression profile
### Import expression matrix

```{r}
min.cells = 30;

fname = '../../input/datasets/HumanBrain/Lake_Fcx_expression_org.RData'
if(!file.exists(fname)) {
  expression = as(readMM('../../input/datasets/HumanBrain/Lake_Fcx_expression.mm'), 'CsparseMatrix')
  gene_names = read.table('../../input/datasets/HumanBrain/Lake_Fcx_gene_annotations.txt', as.is = T, header = T)$HGNC
  sample_annotations = read.table('../../input/datasets/HumanBrain/Lake_Fcx_sample_annotations.txt', header = T)
  
  filtered_cells = which(sample_annotations$Labels %in% c('Endo', 'Per'))
  sample_annotations = sample_annotations[-filtered_cells, ]
  expression = expression[, -filtered_cells]
  dim(expression)
  
  d = Matrix::rowSums(expression > 0);
  gene.filter.mask = d < min.cells;
  sum(gene.filter.mask)
  expression = expression[!gene.filter.mask,]
  gene_names = gene_names[!gene.filter.mask]
  
  dim(expression)
  
  d = Matrix::colSums(expression)
  med = median(d)
  D_inv = Diagonal(n = length(d), x = 1 / d)
  expression  = expression %*% D_inv # normalize
  expression = med*expression # re-scale
  expression = log(1 + expression) # variance-stablize (using log-transform)
  

  expression = as(expression, "sparseMatrix") # convert to dgCMatrix format for passing to RCPPArmadillo (in ACTION)
  
  save(expression, gene_names, sample_annotations, file=fname)
} else {
  load(fname)
}

Labels = sample_annotations$Labels
UL = sort(unique(Labels))
labels = match(Labels, UL)


```


### Run ACTION
```{r}
library(ACTIONet)

reduction.out = reduceGeneExpression(expression, reduced_dim = 30, method = 1, iter = 10)
S_r = reduction.out$S_r
V = reduction.out$V

ACTION.out = runACTION(S_r, k_max = 20, thread_no = 8)

```


# Prune archetypes (new in the latest version of ACTIONet)
```{r}
# Remove non-specific archetypes
reconstruct.out = reconstructArchetypes(expression, ACTION.out$C, ACTION.out$H, -1)

# Remove redundant archetypes
backbone.reduction.out = ACTIONetcore::reduceBackbone(backbone = reconstruct.out$backbone, C_stacked = reconstruct.out$C_stacked, H_stacked = reconstruct.out$H_stacked, z_threshold = 3)

arch.assignments = backbone.reduction.out$representatives
core.archs = sort(unique(arch.assignments))
```

##
```{r}
# Convex-regression using the reduced subset of archetypes
H_final = runsimplexRegression(reduction.out$S_r %*% reconstruct.out$C_stacked[, core.archs], reduction.out$S_r)

# "Impute" gene expression values
metagene.archetype.profile = reduction.out$S_r %*% reconstruct.out$C_stacked[, core.archs]
gene.archetype.profile = V %*% metagene.archetype.profile # Boosts contribution of differential genes
Arch.imputed.profile = gene.archetype.profile %*% H_final # Interpolate expression of cells
rownames(Arch.imputed.profile) = gene_names

dim(Arch.imputed.profile)
```


## Load PCNet (Parsed from: "Systematic Evaluation of Molecular Networks for Discovery of Disease Genes")

```{r}
PCNet = as(readMM('../../input/networks/PCNet/PCNet.mm'), 'CsparseMatrix')
PCNet_genes = read.csv('../../input/networks/PCNet/PCNet_genes.txt', as.is = TRUE)$Genes
rownames(PCNet) = PCNet_genes
colnames(PCNet) = PCNet_genes

```


## Match network and expression profile
```{r}
common_genes = intersect(PCNet_genes, gene_names)

blacklist = c('UBC', 'SUMO1', 'SUMO2', 'SUMO3', gene_names[grep('^RPL|^RPS|^MRP', gene_names)])
# blacklist = c('UBC', 'SUMO1', 'SUMO2', 'SUMO3')
common_genes = setdiff(common_genes, blacklist)

length(common_genes)

order1 = match(common_genes, gene_names)
order2 = match(common_genes, PCNet_genes)

A = as.matrix(Arch.imputed.profile[order1, ])
net = as.matrix(PCNet[order2, order2])

dim(A)
dim(net)
sum(net > 0)/2
```


## Network inference

### Compute summary network per group of cells (i.e., cell types)
```{r}
# load(file = '../../results/Lake_Fcx_mapped_datasets.RData')
require(SCINET)
require(igraph)
require(R.utils)
require(Matrix)

UL = c('Mic') # Only construct networks for Mic for testing
rand_sample_no = 30 # Rather small for testing. Best selected as a proportion of samples in the group
rand_sample_size = 10 # Rather small for test. Suggested: 30

nets = vector("list", length(UL))
names(nets) = UL
for (celltype in UL) {
  printf('Celltype %s\n', celltype)
  
  celltype_samples = which(Labels %in% celltype)
  Net_stats = constructNet_summary(A = A,  net=net, samples = celltype_samples, rand_sample_no = 30, rand_sample_size = 10, thread_no = 8)
  
  # Prune to the most significant edges and construct an igraph object
  Adj = Net_stats$mu
  Adj[Adj < 2 + log10(length(Adj))] = 0
  Adj = as(Adj, 'sparseMatrix');
  specific_net = graph_from_adjacency_matrix(Adj, weighted = TRUE, mode = "undirected")
  V(specific_net)$name = common_genes

  nets[[celltype]] = specific_net
}

G = nets[['Mic']]
EdgeList.Mic1 = as.data.frame(cbind(get.edgelist(G), get.edge.attribute(G, "weight")))
colnames(EdgeList.Mic1) = c('src', 'dst', 'weight')
EdgeList.Mic1 = EdgeList.Mic1[order(EdgeList.Mic1$weight, decreasing = TRUE), ]
```

### Compute a single network per cell
```{r}
# Construct networks for 50 random Mic samples
fullNets = constructNet(A = A, net=net, samples = which(Labels == 'Mic')[1:30], thread_no = 8)
w = rowMeans(fullNets$weights)
subs = fullNets$subs + 1

EdgeList.Mic2 = data.frame(src = common_genes[subs[, 1]], dst = common_genes[subs[, 2]], weight = w)
EdgeList.Mic2 = EdgeList.Mic2[order(EdgeList.Mic2$weight, decreasing = TRUE), ]

```